Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-06-13 content.tex
Revised Examples

We show the usage of \verb;pirouette; by gradually introducing
its features.
First, to get an idea of the baseline error, we measure the Bayesian inference error when we start from a phylogeny generated under a known and standard tree prior.
Second, to establish that the true tree prior is indeed the best, we also measure the inference error made adopting a
best candidate inference model.
Lastly, we quantify the impact of the tree prior in the Bayesian inference. We do so by separating the baseline error and the complete error running the pipeline from a tree generated by an unknown or non-standard tree prior.

@@

In this example we use a standard generative tree prior $\mathit{p_{G}^0}$, namely the Yule (pure-birth) tree prior. 
We choose to use a small tree with six taxa to keep
the calculations short and the figure more readable.
We pick a crown age of ten time units. This value is 
completely arbitrary, but it ties in with the mutation rate 
used in simulating an alignment in the next step.

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny, as described in subsection~\ref{subsec:pipeline}.
In this example, the root sequence consists of four blocks of 250 mononucleotides each, while the per-nucleotide mutation rate is 0.1 mutations per unit time.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A JC site model assumes that mutation rates between nucleotides are equal and constant. 
A strict clock model assumes that the mutation rates 
of all lineages are equal and constant.

@@

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is in fact an inherent part of \verb;pirouette;. 
We call it 'twinning' and we will use it starting from \ref{Comparing to background noise}.

@@

So far we have measured the inference error on a tree
generated according to a known (and standard) tree prior. The goal of \verb;pirouette; is to measure the impact of a novel tree prior.
To do so, in this example, we will use a tree generated by a non-standard tree prior, assumed to be closely related to the Yule tree prior.
To measure the impact of the tree prior, we measure both the baseline error (that serves as a control) and the full inference error: their difference shows the impact of the tree prior.
To do so, we use a twin tree (see Subsection~\ref{subsec:twinning}) for each of the generative and best candidate model.
The research question this example answers is:
"What is the inference error made from a phylogeny, for both a generative model and a best candidate model, compared to the background noise?"

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/errors.png}
  \caption{
    The inference error made 
    for both a generative tree prior and best candidate model
    compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error.
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_3}
\end{figure}
\giovanni{
  I find the figure to be too confused. 
  What if we use instead different plots, 
  and in each of them we just show one inference model's results plus its twin counterpart?
}
\richel{Will do}

@@

We showed how to use \verb;pirouette; to quantify the importance of a tree prior in Bayesian phylogenetics, using the simplest generative tree prior possible.
In principle any other (more complex) generative tree prior can be tested, but we chose to provide the simplest (and fastest to run) examples.
All tests were performed on a single original tree, 
hence not providing us enough statistical power to 
evaluate the level of accuracy of the inference.

However, if the same procedure were repeated and performed on a distribution with a sufficient number of generative trees, 
it would constitute a quantitative and effective assessment of the quality of the inference.
Also a twin tree does not always result in a lower error distribution,
due to the stochasticity in generating the twin tree.

>>>>

We show the usage of \verb;pirouette; by gradually introducing
its features.
First, to get an idea of the baseline error, 
we measure the Bayesian inference error when we start from a phylogeny 
generated under a known and standard tree prior.
Second, to establish that the true tree prior is indeed the best, 
we also measure the inference error made adopting a
best candidate inference model.
Lastly, we quantify the impact of the tree prior in the Bayesian inference. 
We do so by separating the baseline error from the complete error 
running the pipeline starting from a tree generated by an unknown 
or non-standard tree prior.

@@

In this example we use a standard generative tree prior $\mathit{p_{G}^0}$, namely the Yule (pure-birth) tree prior. 
We choose to use a small tree with six taxa, to keep
the calculations short and the figure more readable.
We pick a crown age of ten time units. This value is 
completely arbitrary, but it ties in with the mutation rate 
used in simulating an alignment in the next step.

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny, as described in Subsection~\ref{subsec:pipeline}.
In this example, the root sequence consists of four blocks of 250 
mononucleotides each, while the per-nucleotide mutation rate is 
0.1 mutations per unit time.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A JC site model assumes that mutation rates between nucleotides are equal and 
constant. 
A strict clock model assumes that the mutation rates 
of all lineages are equal and constant.

@@

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is in fact an inherent part of \verb;pirouette;. 
We call it 'twinning' and we will demonstrate it at section
 \ref{Comparing to background noise}.

@@

So far we have measured the inference error on a tree
generated according to a known (and standard) tree prior. 
The goal of \verb;pirouette; is to measure the impact of a novel tree prior.
To do so, in this example, we will use a tree generated by a non-standard 
tree prior, assumed to be closely related to the Yule tree prior.
To measure the impact of the tree prior, we measure both the baseline 
error (that serves as a control) and the full inference error: it is their 
difference that shows the impact of the tree prior.
To do so, we use a twin tree (see Subsection~\ref{subsec:twinning}) for 
each of the generative and best candidate model.
The research question this example answers is:
"What is the inference error made from a phylogeny, 
for both a generative model and a best candidate model, compared to the 
background noise?"

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/errors.png}
  \caption{
    The inference error made 
    for both a generative tree prior and best candidate model
    compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error.
    Vertical dashed lines show the median error value per distribution.
  }
  \label{fig:example_3}
\end{figure}

@@

We showed how to use \verb;pirouette; to quantify the importance of a 
tree prior in Bayesian phylogenetics, using the simplest generative tree 
prior possible.
In principle any other (more complex) generative tree prior can be tested, 
but we chose to provide the simplest (and fastest to run) examples.
All examples used only one original tree,
where any speciation process produces a whole range of trees.
One tree is not enough to determine the impact of a tree prior 
on Bayesian inference.
However, if the same procedure were repeated and performed on a 
distribution with a sufficient number of generative trees, 
it would constitute a quantitative and effective assessment of the 
quality of the inference.
Also a twin tree does not always result in a lower error distribution,
due to the stochasticity in generating the twin tree.
